\documentclass{report}

\usepackage{hyperref}
\usepackage{color}
\usepackage[usenames]{xcolor}
\usepackage{tikz}

\usepackage{algorithm}
\usepackage{algpseudocode}

% New definitions
\algnewcommand\algorithmicswitch{\textbf{case}}
\algnewcommand\algorithmiccase{\textbf{}}
% New "environments"
\algdef{SE}[SWITCH]{Switch}{EndSwitch}[1]{\algorithmicswitch\ #1\ {\bf of}}{\algorithmicend\ \algorithmicswitch}%
\algdef{SE}[CASE]{Case}{EndCase}[1]{\algorithmiccase\ #1 :}{\algorithmicend\ \algorithmiccase}%
\algtext*{EndSwitch}%
\algtext*{EndCase}%

%\usepackage[T1]{fontenc}

\definecolor{lgray}{gray}{0.9}

\usepackage{listings}

\lstnewenvironment{tsllisting}[1]
{\vspace{3mm}
 \lstset{
    backgroundcolor=\color{lgray},
    basicstyle=\small\ttfamily, 
%    keywordstyle=\bfseries,
    keywordstyle=\underbar,
    identifierstyle=,
    commentstyle=\slshape,
    stringstyle=,
    showstringspaces=false,
    keywords={after,
              always,
              assert,
              assign,
              assume, 
              before,
              bool,
              case,
              choice, 
              cond,
              const,
              controllable, 
              default,
              derive,
              do, 
              else, 
              endtemplate,
              enum,
              export,
              false,
              forever, 
              fork, 
              function, 
              goal,
              if, 
              import,
              init,
              invisible, 
              mem,
              out,
              pause,
              break,
              post, 
              procedure, 
              process,
              return,
              sint,
              stop, 
              struct,
              task, 
              template, 
              true,
              typedef,
              uint,
              uncontrollable, 
              using, 
              void,
              while,
              wait},
    sensitive=false,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    numberstyle=\tiny,
    stepnumber=1,
    numbersep=1pt,
    emphstyle=\bfseries,
    belowskip=0pt,
    aboveskip=0pt,
    #1
}}{\vspace{3mm}}

\lstnewenvironment{bnflisting}[1]
{\vspace{3mm}
 \lstset{
    backgroundcolor=\color{lgray},
    basicstyle=\small\ttfamily,
    keywordstyle=\underbar,
    identifierstyle=,
    commentstyle=\slshape,
    stringstyle=,
    showstringspaces=false,
    keywords=,
    sensitive=false,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    numberstyle=\tiny,
    stepnumber=1,
    numbersep=1pt,
    emphstyle=\bfseries,
    belowskip=0pt,
    aboveskip=0pt,
    #1
}}{\vspace{3mm}}


\newcommand{\src}[1]{\texttt{#1}}

\newcommand{\tsl}{TSL2 }

\newcommand{\comment}[1]{{\textit{\textbf{#1}}}}


\author{Leonid Ryzhyk}

\title{\tsl Reference Manual}

\begin{document}

\maketitle

\tableofcontents

\chapter{Overview}

\section{Static and dynamic namespaces}\label{s:o:namespace}

\tsl supports two namespaces: the \emph{static namespace} and the 
\emph{dynamic namespace}.  The static namespace is populated with 
compile-time objects: \emph{types} and \emph{constants}.  The 
dynamic namespace is populated with runtime objects: 
\emph{processes}, \emph{variables}, \emph{methods}, and 
\emph{wires}.  Static objects are uniquely identified by their 
name and syntactic scope.  In contrast, runtime objects can be 
instantiated multiple times within the specification and hence 
must be referred to relative to their runtime scope.  

\section{Templates}\label{s:o:templates}

Templates are the principal mechanism for managing both static and 
dynamic namespaces.  A template models an entity, such as a 
device, an OS, or a device driver.  It declares a set of static 
objects (types and constants) and a set of runtime objects.  
Static objects declared inside a template can be referenced from 
any part of the specification via the template name.  Runtime 
objects are instantiated together with the template and can be 
accessed via a reference to a template \emph{instance}.

The following template declares type \src{word} (static object) 
and variable \src{x} (runtime object):
\begin{tsllisting}{}
template A
  // type declaration
  typedef uint<16> word;
  // variable declaration
  export word x;
endtemplate
\end{tsllisting}
The \src{word} type is globally visible via the \src{::}-notation 
as \src{A::word}.  It can be used even if template \src{A} is 
never instantiated.  In contrast, variable \src{x} can only be 
accessed via an instance of \src{A}.  

A template is instantiated inside another template.  The only 
exception is the \src{main} template, which is implicitly 
instantiated in the top-level scope.  Every complete \tsl 
specification must contain a template called \src{main}.  

In the following example, the \src{main} template creates an 
instance of \src{A}, making its variables accessible from 
\src{main} via instance name:
\begin{tsllisting}{}
template main
  instance A a;

  process pmain {
    // assigning variable x of template instance a.
    a.x = 16'd0;
  }
endtemplate
\end{tsllisting}

The template instantiation mechanism gives rise to an 
\emph{instance tree} with the \src{main} template as its root.  
Dynamic objects within the tree are accessed using hierarchical 
identifiers such as \src{a.x}.  This mechanism allows accessing 
objects down the branch of the instance tree, starting at the 
local template.  In practice, it is often necessary to access 
objects instantiated in other parts of the tree.  This is achieved 
in a structured way using \emph{template ports}.  

A template port is an alias to a template of a given type bound at 
the time of instantiation.  For example, template \src{B} below 
declares port \src{aa} of type \src{A}, which makes runtime 
objects in the scope of \src{A} visible from \src{B}.  Both 
templates are then instantiated in the \src{main} template, with 
the instance of \src{A} connected to port \src{aa} of \src{B}.
\begin{tsllisting}{}
// template B with port aa of type A
template B(A aa)
  process proc {
    aa.x = 16'd0;
  };
endtemplate

template main
  // create instances of A and B; connect port aa of
  // B to the instance of A.
  instance A a;
  instance B b(a);
endtemplate
\end{tsllisting}

The following diagram illustrates the resulting instance tree and 
the link between different branches of the tree via port \src{aa}.

\begin{tikzpicture}
    \node[rectangle] (m) at (0,1) {main};
    \node[rectangle] (a) at (-1,0) {main.a};
    \node[rectangle] (b) at (1,0) {main.b};
    \path[->] (m) edge (a);
    \path[->] (m) edge (b);
    \path[->,dashed] (b) edge node[auto]{aa} (a);
\end{tikzpicture}

Another mechanism for managing name spaces is \emph{template 
inheritance}.  Using inheritance, one can create generic templates 
that capture common properties of a family of entities, leaving 
some of the properties underspecified.  The generic template can 
be specialised by a child template that fills in the missing 
details.  For example, the following template models common 
device-class callbacks that must be implemented by any OS 
specification for the IDE device class.  Note that callbacks are 
defined without bodies, as the exact behaviour is OS-specific.
\begin{tsllisting}{}
template ide_os
  procedure void write_sectors(uint<48> lba, 
    uint<16> sectors, uint<32> buf, bool xfer_error);
  procedure void read_sectors(uint<48> lba, 
    uint<16> sectors, uint<32> buf, bool xfer_error);
  procedure void reset();
  ...
endtemplate
\end{tsllisting}
This template is specialised by the \src{l4\_ide\_os} template,
that describes the IDE driver interface defined by the seL4 OS.
\begin{tsllisting}{}
template l4_ide_os(l4_ide_drv drv)
  // the derive statement is used to establish the 
  // inheritance relation
  derive ide_os;

  // additional specification items can be declared in the 
  // child template
  export iostatus reset_status = ionone;

  // the child template implements methods inherited from 
  // the parent.
  procedure void write_sectors(uint<48> lba, 
    uint<16> sectors, uint<32> buf, bool xfer_error)
  {
    assert (lba == r_lba);
    ...
  }
\end{tsllisting}

A template that has part of its functionality, namely method body 
or wire assignment, underspecified is called \emph{pure template}.  
Pure templates can be used in derive statements and in port 
declarations, however they cannot be instantiated.

Template inheritance is subject to the following rules:
\begin{itemize}
    \item Derived template must re-declare all ports of all its 
        parent templates with the same names and types.  It can 
        also declare additional ports not found in any of its 
        parents.

    \item If a template has multiple parent templates, then the 
        namespaces of parent templates (consisting of variable, 
        wire, method, goal, and process names) must not overlap.

    \item If the derived template re-declares a wire or method 
        declared in one of its parents, the new declaration must 
        have the same signature as the parent declaration.

    \item The child template can override wire and method 
        declarations of its parent templates.  Other parts of the 
        parent template, including variable, goal, and process 
        declarations, are inherited by the child template and 
        cannot be overriden.
\end{itemize}



\section{Execution model}\label{s:o:execmodel}

All state transitions in \tsl occur in the context of 
\emph{processes}.  Multiple processes can be enabled at the same 
time; however exactly one process participates in each individual 
transition.  Process transitions are atomic with respect to other 
processes.  Process that perform the next transition is chosen by 
the scheduler among enabled processes.  Scheduling is fair, i.e., 
if a process stays enabled sufficiently long, it will eventually 
get scheduled.

Processes are declared inside templates using the \src{process} 
keyword.  Such processes become runnable in the initial state of 
the specification.  Additional processes can be spawned at runtime 
using the \src{fork} construct.  Syntactic constraints of the 
language (namely, the lack of support for recursion) guarantee 
that only a bounded number of processes can be spawned at any 
time.  

In the following example, template \src{A} contains a single 
static process \src{psndrcv}, which spawns two subprocesses 
\src{psend} and \src{preceive}.
\begin{tsllisting}{}
template A

process psndrcv {
  fork {
    psend:    forever send();
    preceive: forever receive();
  };
  shutdown();
};
...

endtemplate
\end{tsllisting}

%In addition to statically and dynamically created processes, every 
%\tsl specification contains an implicit \emph{idle} 

A process state transition starts in the current program location 
and stops at the next \emph{pause location}.  Pause locations can 
be explicit or implicit.  Explicit pause locations are declared 
using \src{pause}, \src{wait}, and \src{stop} constructs.  
Implicit pause locations are introduced automatically by the 
compiler in the following cases:

\begin{itemize}
    \item Before \src{fork} statements
    \item Before magic blocks~\ref{s:o:magic}
    \item On entries to and exits from controllable and 
        uncontrollable tasks (see below)
\end{itemize}

Process behaviour can be factored into \emph{methods}.  A process 
can invoke methods declared inside its own template as well as in 
other template instances, via hierarchical identifiers discussed 
in the previous section.  There are three types of methods in 
\tsl: \emph{functions}, \emph{procedures}, and \emph{tasks}.  
Functions and procedures must complete instantaneously, i.e., they 
are not allowed to contain explicit or implicit pauses.  In 
addition, functions are not allowed to have side effects, i.e., 
they cannot modify any global variables, perform pointer 
dereferences, or contain assertions (pointer dereferences and 
assert statements can have the side effect of taking the system 
into an error state).  

Tasks can take time to execute.  A task can have an optional 
\src{controllable} or \src{uncontrollable} qualifier.  
Controllable tasks are the only kind of task that can be invoked 
from synthesised code; although they can also be called from 
manually written code.  They are used to model the device register 
interface and OS callbacks.  In the current implementation of the 
\tsl compiler, controllable tasks are subject to additional 
constraint: they are not allowed to contain internal pause 
locations, i.e., a controllable task must always complete in a 
single transition. 

Uncontrollable tasks represent driver methods invoked by the OS or 
the device.  Uncontrollable tasks are the only kind of task that 
can contain automatically generated code, i.e., a magic block can 
only be placed inside an uncontrollable task.  

Controllable and uncontrollable task invocations introduce 
implicit pause locations on entry and return from the task.  In 
the following example execution of process \src{p1} will consist 
of three transitions.
\begin{tsllisting}{}
template A
  uint<16> x;

  process p1 {
    x=1;
    t1(x);
  };

  task uncontrollable void t1(uint<16> arg) {};
endtemplate
\end{tsllisting}
The first transition assigns global variable \src{x} and sets the 
argument of task \src{t1} to be equal to \src{x}.   An implicit 
pause location is inserted at this point.  The second transition 
executes the body of task \src{t1}, which completes 
instantaneously, as it does not contain any pause locations.  
Another implicit pause is inserted at the task return location.  
The remainder of the process is executed in the third transition.  
Note that other processes can run and potentially modify variable 
values in between transitions 1 and 2 and transitions 2 and 3.

Tasks without \src{controllable} and \src{uncontrollable} 
qualifiers behave as if the body of the task was inlined at the 
call location.  No additional pauses are inserted before or after 
the task.

\section{Variables and wires}\label{s:o:variables}

\emph{Variables} are used to store state that persists for the 
lifetime of the variable.  Variables in \tsl can be declared in 
the template, process, or method scope.  Template-scope variables, 
also called \emph{global variables}, are instantiated together 
with the template and are visible from anywhere inside the 
template.  In addition, variables declared with the \src{export} 
qualifier can be accessed from other templates via their 
hierarchical identifiers.  Process and method variables are only 
visible within the syntactic scope of the process or method where 
they are declared.

In contrast to variables, \emph{wires} are simply aliases to 
expressions defined over global variables and are not allocated 
their own storage.  Wires keep their values throughout a 
transition and are updated at the end of the transition.  Initial 
value of a wire is computed based on initial values of global 
variables.

\begin{tsllisting}{}
template A
  uint<16> x = 0;
  wire uint<16> w = x + 1;
endtemplate
\end{tsllisting}


\section{Correctness specifications}\label{s:o:correctness}

\tsl provides several mechanisms for specifying correctness 
conditions over system behaviour:
\begin{itemize}
    \item \emph{Goals.} A goal is a side-effect-free boolean 
        expression over global variables that must hold infinitely 
        often in any infinite run of the system.  A template can 
        declare any number of goals.  In addition, \tsl defines 
        two implicit goals.  The first one requires the system to 
        be outside of a magic block infinitely often; in other 
        words, the system cannot stay inside a magic block 
        forever.  The second implicit goal requires that the 
        system never enters an \emph{error state} (see below).
    \item \emph{Assertions.}  An \src{assert} statement can be 
        placed anywhere inside processes, tasks, and procedures.  
        It defines a condition whose violation immediately 
        transitions the system to an error state.
    \item \emph{Postconditions}.  Magic block postconditions are 
        discussed in Section~\ref{s:o:magic} above.
\end{itemize}



\section{Magic blocks}\label{s:o:magic}

A \emph{magic block} is a place holder for automatically generated 
code.  \tsl supports two types of magic blocks.  Magic blocks with 
\emph{postconditions} specify a condition that must hold upon 
returning from the magic block.  

%Note that for synthesis to succeed the resulting system must 
%satisfy all correctness conditions described in the following 
%section, and not just magic block postconditions.

\begin{tsllisting}{}
template main
  task uncontrollable bool probe() {
    // Magic block with postcondition
    {...} post((os.reset_status == iosuccess) ||
               (os.reset_status == ioerror));
    if (os.reset_status == iosuccess) {
      return true;
    } else {
      return false;
    };
};
endtemplate
\end{tsllisting}

The second type of magic blocks are magic blocks with 
\emph{goals}.  These are needed because not all correctness 
conditions can be captured using postconditions and assertions.  
For correctness conditions specified using goals (see 
Section~\ref{s:o:correctness}), the synthesis algorithm computes a 
strategy for each goal; however the scheduling of strategies is 
left to the user, i.e., the user decides when to execute each 
strategy by assigning goals to magic blocks. \comment{TODO: I 
still have not figured out the exact meaning of this construct.}

%\section{Constraints on the environment}

\chapter{Syntax reference}

\section{Literals}\label{s:r:literals}

\tsl supports boolean literals ``\src{true}'' and ``\src{false}'' 
and Verilog-style binary, octal, decimal and hexadecimal integer 
literals.  The exact number of bits can be specified for each 
integer literal.  

\begin{bnflisting}{}
<intLit> := <decNumber>
          | [<width>] "'b"  <binNumber>
          | [<width>] "'sb" <binNumber>
          | [<width>] "'o"  <octNumber>
          | [<width>] "'so" <octNumber>
          | [<width>] "'d"  <decNumber>
          | [<width>] "'sd" <signedDecNumber>
          | [<width>] "'h"  <hexNumber>
          | [<width>] "'sh" <hexNumber>
<width> := <decNumber>
\end{bnflisting}

Examples of integer literals:
\begin{tsllisting}{}
uint<8> x;
sint<8> y;

x = 255;
x = 8'd35;
x = 8'b01010101;
x = 8'b1;

y = 8'sb11111111;
y = 8'sd-6;
\end{tsllisting}

In case literal width is not specified explicitly, the compiler 
assumes width sufficient to encode the given integer value, for 
example literal $5$ is assumed to be $3$ bits wide, as $3$ bits 
are sufficient to encode values between $0$ and $5$.  The compiler 
does not perform automatic truncation or extension of integers.  
For example, the following assignment statement is invalid, as the 
left and the right-hand sides of assignment have different width:

\begin{tsllisting}{}
uint<8> x = 0; // error: x has width 8, while
               // literal 0 has widt 1

uint<8> y = 8'd0; // ok
\end{tsllisting}


\section{Identifiers}\label{s:r:identifiers}

Identifiers are used throughout \tsl specification to refer to 
static and runtime objects (Section~\ref{s:o:namespace}).  \tsl 
supports three forms of identifiers:

\paragraph{Simple identifiers.}  A simple identifier is a name of 
a static or runtime object visible within the current syntactic 
scope.  This includes objects declared within the current scope or 
in one of its parent scopes.  For example, simple identifier 
\src{x} used in a method body can refer to a local variable or 
argument of the method, template-global variable, or a type or 
constant declared in the template or top-level scope.

\begin{bnflisting}{}
<ident> := (<letter> | "_") (<letter> | <digit> | "_")*
\end{bnflisting}

\paragraph{Static scoped identifiers.} These identifiers refer to 
static objects declared within template scope.  They can be used 
to refer to static objects declared in templates other than the 
local template.

\begin{bnflisting}{}
<staticIdent> := <ident> "::" <ident>
\end{bnflisting}

\paragraph{Hierarchical identifiers.}  These identifiers can only 
be used to refer to runtime objects.  A dynamic identifier is a 
dot-separated sequence of literals that traverses the instance 
tree (Section~\ref{s:o:templates}) via port and instance names:

\begin{bnflisting}{}
<dynIdent> := <ident> ["." <ident>]*
\end{bnflisting}

The following example illustrates the use of different types of 
identifiers.

\begin{tsllisting}{}
template A
  typedef uint<16> word;
  export word x;
endtemplate
    
template B(A aa)
  // Reference to the word type declared in template A
  // via static scoped identifier;
  A::word y;

  process proc {
    // In the following statement:
    // * y is a simple identifier that refers to a variable
    //   declared in the local template
    // * aa.x is a hierarchical identifier that refers to
    //   variable x declared in template A.
    y = aa.x;
  };
endtemplate

template main
  // create instances of A and B; connect port aa of
  // B to the instance of A.
  instance A a;
  instance B b(a);
endtemplate
\end{tsllisting}


\section{Types}

Type expressions can occur in various contexts in a \tsl 
specification, including variables, method, wire, type, and 
constant declarations.  The language currently supports arbitrary 
fixed-width signed and unsigned integers, booleans, enums, 
structs, arrays, and pointers.

\begin{bnflisting}{}
<typeSpec>   := ( <sintType>    // signed int
                | <uintType>    // unsigned int
                | <boolType>    // boolean
                | <userType>    // user-defined type name
                | <enumType>    // enumeration
                | <structType>) // struct
                <typeModifier>* // type modifiers

<sintType>   := "sint" "<" <decimalNumber> ">"
<uintType>   := "uint" "<" <decimalNumber> ">"
<boolType>   := "bool"
<userType>   := <staticIdent>
<enumType>   := "enum" "{" (<ident> ",")*  <ident> "}"
<structType> := "struct" "{" (<typeSpec> <ident> ";")+ "}"

// A type modifier is either array dimension or the pointer
// modifier ("*")
<typeModifier> := ("[" <expr> "]")
                | "*"
\end{bnflisting}

\tsl enum's are different from C-style enum's in that they are not 
integers.  In particular, a variable of enum type cannot be cast 
to an integer.  Such a variable can only take one of the values in 
the enumeration and not any other arbitrary integer value.  
Furthermore, an enum declaration cannot assign integer values to 
enumerators.

Type expressions are subject to the following constraints:
\begin{itemize}
    \item Enumerations can only be declared in \src{typedef}
        statements, e.g., they cannot be used in variable or 
        method declarations.  This ensure that every enum type has 
        a name.
    \item Variable-size arrays are not supported: array dimensions 
        must be compile-time constants.
\end{itemize}

Examples of type expressions:

\begin{tsllisting}{}
typedef uint<16> t1;
typedef sint<13> t2;
typedef enum {e1, e2, e3} t3;
typedef struct {t1 f1; t2 f2;} t4;
const t1 c1 = 16'd5;
typedef t4[c1] arrtype;
typedef struct {bool f1; uint<16>[10] f2;} * structptr;
\end{tsllisting}

\section{Expressions}

\tsl expressions are constructed from identifiers 
(Section~\ref{s:r:identifiers}), literals
(Section~\ref{s:r:literals}), and method 
invocations~\ref{s:r:invocation} using operators summarised in 
Table~\ref{t:ops} in the order of decreasing precedence.

\begin{table}
\begin{small}
\noindent
\begin{tabular}{|l|l|l|l|p{0.3\linewidth}|}
    \hline
    {\bf operator}                   & {\bf syntax} & {\bf arg type} & {\bf res type} & {\bf comment} \\
    \hline
    \hline 
    {\tt[:]}                         & {\tt e[l:h]} & integer        & unsigned int   & bit slice \\
    {\tt[~]}                         & {\tt e[i]}   & array          & -              & array index \\
    {\tt.}                           & {\tt e.f}    & struct         & -              & struct field \\
    {\tt\verb=->=}                   & {\tt\verb=e->f=}  & struct pointer & -              & struct dereference \\
    \hline
    \src{!}                          & prefix  & bool           & bool           & boolean negation \\
    {\tt\verb=~=}                    & prefix  & integer        & integer        & bit-wise negation \\
    {\tt-}                           & prefix  & integer        & integer        & unary minus \\
    {\tt*}                           & prefix  & pointer        & -              & pointer dereference \\
    {\tt \&}                         & prefix  & any            & pointer        & address-of \\
    \hline
    {\tt ==, !=}                     & infix   & any            & bool           & \\
    {\tt\verb#<, <=, >, >=#}         & infix   & integer        & bool           & \\
    \hline    
    {\tt\&}                          & infix   & integer        & integer        & bit-wise and \\
    \hline    
    {\tt |}                          & infix   & integer        & integer        & bit-wise or \\
    \hline    
    {\tt\verb#^#}                    & infix   & integer        & integer        & bit-wise xor \\
    \hline    
    {\tt\&\&}                        & infix   & bool           & bool           & boolean and \\
    \hline    
    {\tt||}                          & infix   & bool           & bool           & boolean or \\
    \hline    
    {\tt\verb#=>#}                   & infix   & bool           & bool           & boolean implication\\
    \hline    
    {\tt*}                           & infix   & integer        & integer        & multiplication \\
    \hline    
    {\tt\%}                          & infix   & integer        & integer        & residue \\
    \hline    
    {\tt+}                           & infix   & integer        & integer        & plus \\
    \hline    
    {\tt-}                           & infix   & integer        & integer        & minus \\
    \hline
\end{tabular}
\end{small}
\caption{\tsl operators}\label{t:ops}
\end{table}

In addition, \tsl supports three kinds of conditional expressions: 
\src{if-else} expressions (or ternary expressions), 
\src{case}-expressions, and \src{cond}-expressions.  

\begin{bnflisting}{}
<ternExp> := "if" <expr> <expr> "else" <expr>
<caseExp> := "case" "(" <expr> ")" "{" (<expr> ":" <expr> ";")* 
                 (<expr> ":" <expr> ";")* 
                 ["default" ":" <expr> ";"]
            "}"
<condExp> := "cond" "{"
                  (<expr>    ":" <expr> ";")*
                  ["default" ":" <expr> ";"]
              "}"
\end{bnflisting}

A \src{case}-expression chooses a value based on the value of its 
key expression, whereas a \src{cond}-expression chooses a value to 
return by evaluating a series of conditions in order.  Note that 
\src{if-else} expressions are different from \src{if} statements 
described in Section~\ref{s:r:coniditional}.

Finally, \tsl supports struct expressions with explicit or implicit 
field names:

\begin{bnflisting}{}
<structExp>   := <staticIdent> "{" 
                     (<namedFields> | <anonFields>) 
                 "}"
<namedFields> := "." <ident> "=" <expr> 
                 [("," "." <ident> "=" <expr>)*]
<anonFields> := <expr> 
                 [("," <expr>)*]
\end{bnflisting}

Note that \tsl does not provide type casting operations.  In 
particular, it is impossible to convert a pointer to an integer or 
an integer to a pointer.  In addition, the \tsl compiler enforces the 
following type and memory safety rules:
\begin{itemize}
    \item No arithmetic operations are allowed on pointers
    \item Bit-wise operators must be applied to integer operands 
        of the same width
    \item \src{==} and \src{!=} operations can only be applied to 
        operators of identical types (including width and 
        signedness)
    \item Labels in a \src{case}-expression and conditions in a 
        \src{cond}-expression must be side-effect free
    \item All branches of a \src{case}, \src{cond}, or \src{if-else}
        expression must return values of the same type
    \item Lower and upper bounds of a bit slice must be constant 
        expressions, lower bound must be less than or equal to the 
        upper bound, and both bound must be smaller than the width
        of the argument.
    \item The address-of operator (\src{\&}) can only be applied 
        to memory variables (see Section~\ref{s:r:localvar}).
\end{itemize}

\subsection{Non-deterministic expression}

Special expression \src{*} is used to generate non-deterministic 
values of arbitrary type.  The exact type is derived from the 
context, as illustrated in this example:
\begin{tsllisting}{}
x = *;  // Generate random value of the same type as x.
f(*,*); // Generate random values that match argument 
        // types of f.
\end{tsllisting}
Note that \src{*} cannot be used as an atom in a bigger expression,
for example, the following is not valid:
\begin{tsllisting}{}
x = * + 2; // Invalid use of *
\end{tsllisting}

\subsection{L-expressions}\label{s:r:lexpr}

\emph{L-expressions} are a subset of \tsl expressions that can be 
used as the left-hand side of assignments statement and as output 
arguments of methods.  L-expressions are defined by the following 
rules:
\begin{enumerate}
    \item Names of variables visible in the current scope, 
        including global variables, local variables, and function 
        arguments are L-expressions
    \item All valid expressions of the form \src{*e} and 
        \src{e->f} are L-expressions.
    \item If \src{e} is an L-expression then the following are 
        also L-expressions:
        \begin{itemize}
            \item \src{e.f}
            \item \src{e[i]}
            \item \src{e[l:h]}
        \end{itemize}
\end{enumerate}

\subsection{Expression evaluation}

Expression evaluation order matters in cases when expression 
operands contain method calls, which may have side effects.  
\comment{The following description reflects how the compiler works 
at the moment; however this behaviour is counter-intuitive and 
should be fixed to be more C-like}.  All expression operands are 
evaluated before the expression itself is evaluated.  In 
particular, if the expression contains method calls, then all of 
these calls are performed before the expression is evaluated.

\section{Statements}

Statements are used to describe process behaviour and are used in 
process and method bodies and always-blocks 
(Section~\ref{s:r:always}).

\begin{bnflisting}{}
<statement> := <varDecl>  // local var declaration
             | <sseq>     // sequential block
             | <spar>     // parallel blocks
             | <sforever> // forever loop
             | <sdo>      // do loop
             | <swhile>   // while loop
             | <sfor>     // for loop
             | <sbreak>   // break out of a loop
             | <schoice>  // nondeterministic choice
             | <spause>   // pause
             | <swait>    // wait on a condition
             | <sstop>    // stop
             | <sassert>  // assertion
             | <sassume>  // assumption
             | <site>     // if-then-else statement
             | <scase>    // case statement
             | <sinvoke>  // method invocation
             | <sassign>  // variable assignment
             | <sreturn>  // return statement
             | <smagic>   // magic block
\end{bnflisting}

\subsection{Variable declarations}\label{s:r:localvar}

Variable declaration statements are used to declare local 
variables visible within the syntactic scope of a process, method 
or always-block.  A local variable is visible everywhere within 
the process, task, or always-block regardless of the exact 
location where it has been declared.  A variable declaration 
consists of an optional \src{mem} qualifier, variable type and 
name and optional initial assignment:
\begin{bnflisting}{}
<varDecl> := ["mem"] <typeSpec> <ident> ["=" <expr>]
\end{bnflisting}
The \src{mem} qualifier labels the variable as in-memory variable,
which allows the address-of operator to be applied to this 
variable or any of its fields.  In addition, in performing pointer 
analysis of a \tsl specification, one can assume that a pointer 
can only point to an in-memory variable of a matching type.

In case variable declaration does not specify an initial value for 
the variable, the value for the variable is chosen 
non-deterministically.

\subsection{Sequential blocks} 

C-style sequential blocks; nothing fancy here:
\begin{bnflisting}{}
<sseq> = "{" (<statement> ";")* "}"
\end{bnflisting}

\subsection{Parallel blocks}\label{s:r:fork}

The \src{fork} construct is used to spawn several child processes.  
\begin{bnflisting}{}
<spar> := "fork" "{" 
               (<ident> ":" <statement> ";")*
          "}"
\end{bnflisting}
Each child process is assigned a unique label, which helps 
identify processes during debugging.  The spawning process blocks 
waiting for all forked processes to terminate.  Forked processes 
terminate when all of them have reached a \emph{final} state 
(i.e., a process cannot terminate without waiting for its siblings 
to be in final states as well).  A process is in a final state 
when it reaches either the end of its execution (i.e., executes 
its last instruction) or a \src{stop} statement 
(Section~\ref{s:r:pause}).  

In the following example two dynamically spawned processes send 
and receive data in an infinite loop.
\begin{tsllisting}{}
process psndrcv {
  fork {
    psend:    forever {
                  stop;   // final state
                  send();
              };
    preceive: forever {
                  stop;   // final state
                  receive();
              };
  };
  shutdown();
};
\end{tsllisting}
At each iteration of the loop, they go through final states by 
\src{stop}.  When both processes are in their respective final 
states, the entire \src{fork} block \emph{may} terminate, with the 
parent process moving on to the next statement; however, it is not 
required to terminate and can continue executing \src{forever} 
loops.  The choice between terminating and continuing execution is 
performed nondeterministically by the environment.  In contrast, 
in the following example, once both processes reach their final 
control locations, the fork block terminates instantaneously, as 
neither process can execute any more transitions:
\begin{tsllisting}{}
process psndrcv {
  fork {
    psend:    send();
    preceive: receive();
  };
  shutdown();
};
\end{tsllisting}

\subsection{Loops}  

In addition to conventional do, while, and for loops, \tsl 
supports forever loops, which are equivalent to \src{while(true)}.
\begin{bnflisting}{}
<sdo>      := "do" <statement> "while" "(" <expr> ")"
<swhile>   := "while" "(" <expr> ")" <statement>
<sfor>     := "for" "(" [<statement>] ";" 
                        <expr> ";" 
                        <statement> ")" 
                  <statement>
<sforever> := "forever" <statement>
\end{bnflisting}

At the moment \tsl does not allow \emph{instantaneous loops}, 
i.e., every possible path through the loop body must contain an 
explicit or implicit pause location (Section~\ref{s:o:execmodel}).  
This restriction is enforced by the compiler via a simple static 
check.  Note that there is no implicit pause before the body of the 
loop, i.e., execution enters the loop instantaneously and 
continues until reaching a pause location inside the loop.  For 
example, in the following example, process \src{foo} executes two 
transitions: (1) \src{x=16'd0; (x<1) == true; x=x+16'd1}, and 
(2) \src{(x<1) == false; y=16'd0}.
\begin{tsllisting}{}
process foo {
    x = 16'd0;
    while (x < 1) {
        x = x + 16'd1;
        pause;
    };
    y = 16'd0;
};
\end{tsllisting}

The \src{break} statement can be used anywhere inside a loop to 
transfer control to the first instruction following the body of 
the innermost.

\subsection{Non-deterministic choice}  The \src{choice} construct 
allows the environment to non-deterministically choose between
two or more actions:
\begin{bnflisting}{}
<schoice> := "choice" "{"
               (<statement> ";")*
             "}"
\end{bnflisting}

\subsection{Pause statements}\label{s:r:pause}

\tsl offers three ways to insert an explicit pause location 
(Section~\ref{s:o:execmodel}) in the control flow of a process or 
method: \src{wait}, \src{pause}, and \src{stop}
statements.
\begin{bnflisting}{}
<swait>  := "wait" "(" <expr> ")"
<spause> := "pause"
<sstop>  := "stop"
\end{bnflisting}

The \src{wait} statement inserts a pause location and disables the 
current process until the wait condition becomes true.  If the 
wait condition stays true for sufficiently long time, the process 
is guaranteed to eventually leave the pause location.  Note that 
if the wait condition is already true when the process enters the 
pause location, the current transition terminates anyway.

The \src{pause} statement is a shortcut for \src{wait(true)}.  
\src{stop} behaves like \src{pause}, but additionally marks the 
current process state as final (Section~\ref{s:r:fork}).

\subsection{Assertions} 

An assertion behaves as a no-op if its argument evaluates to true, 
and causes a transition to an error
state otherwise.
\begin{bnflisting}{}
<sassert> := "assert" "(" <expr> ")"
\end{bnflisting}
The argument of an \src{assert} statement must be a 
side-effect-free expression.

\subsection{Assumptions}  

\begin{bnflisting}{}
<sassume> := "assume" "(" <expr> ")"
\end{bnflisting}

Assumptions are used to constrain possible system behaviours.  
Similar to assertions, they specify constraint that must hold for 
any valid execution of the system.  However, while an assertion 
causes transition to an error state if its condition is violated, 
the \src{assume} statement prunes all transitions that violate the 
assumption.

Assumptions are particularly useful in imposing restrictions on 
randomly generated values, as illustrated by the following 
example.
\begin{tsllisting}{}
// assign random values for stopbits and data vars
stopbits  = *;
data      = *;
// only certain combinations of stopbits and data values 
// are valid
assume(((stopbits==UART_STOP_BITS_15) && (data==4'd5)) || 
        ((stopbits==UART_STOP_BITS_2) && (data!=4'd5)));
case (data) {
    4'd5:       data_bits = CUART_DATA5;
    4'd6:       data_bits = CUART_DATA6;
    4'd7:       data_bits = CUART_DATA7;
    4'd8:       data_bits = CUART_DATA8;
    default: assume(false); // other values are not allowed
};
\end{tsllisting}

\subsection{Conditional statements}\label{s:r:coniditional}

\tsl supports C-style \src{if-else} statements and \src{case} 
statements:
\begin{bnflisting}{}
<site>  := "if" "(" <expr> ")" <statement>
             ["else" <statement>]
<scase> := "case" "(" <expr> ")" "{"
             (<expr> ":" <statement> ";")*
             ["default" ":" <statement> ";"
           "}"
\end{bnflisting}

Labels of the \src{case} statement can be arbitrary 
side-effect-free deterministic expressions of a matching type.  At 
runtime, the first matching label is selected.  At most one branch 
of a case statement is executed; execution does not fall through 
to the next label automatically.  Unlike in C, the \src{break} 
statement cannot be used to break out of a case clause.  While 
\src{break} is allowed inside the body of a \src{case}, it has the 
effect of breaking out of the innermost loop.

\subsection{Method invocation}\label{s:r:invocation}

Method invocations can occur as atoms in expressions as well as 
standalone statements.  Method name is a hierarchical identifier 
that refers either to a method declared in the local template or 
exported from another template.
\begin{bnflisting}{}
<sinvoke> := <dynIdent> "(" [<expr> [("," <expr>)*]] ")"
\end{bnflisting}
Method arguments must match the number and types of formal 
arguments in the method declaration.  Output 
arguments~\ref{s:r:method} must be L-expressions~\ref{s:r:lexpr}.

\subsection{Assignment statements}

\begin{bnflisting}{}
<sassign> := <expr> "=" <expr>
\end{bnflisting}

The left-hand side of an assignment statement must be an 
L-expression (Section~\ref{s:r:lexpr}).  Types of left- 
and right-hand side expressions must match, including sign 
and width for integer types.

\subsection{Return statements}\label{s:r:return}

\begin{bnflisting}{}
<sreturn> := "return" [<expr>]
\end{bnflisting}

Return statements are only allowed in method bodies.  If the 
method is a \src{void} method, return must not have an argument; 
otherwise it must have an argument whose type matches the return 
type of the method.  Every path through a non-void method must end 
with a \src{return} statement.  Statements following a 
\src{return} statement are ignored and control transfer to the 
location immediately following the call site.  

If the method is a controllable or uncontrollable task then an 
implicit pause location (Section~\ref{s:o:execmodel}) is 
introduced at the return location.  In this case, if the value 
returned by the task is used as the right-hand side of an 
assignment, the value of the left-hand side is modified at the 
beginning of the next transition performed by the process.

\subsection{Magic blocks}

\begin{bnflisting}{}
<smagic> := "{" "..." "}" ("using" <ident> | "post" <detexpr>)
\end{bnflisting}

See section~\ref{s:o:magic}.

\section{\tsl file structure}

A \tsl file consists of \emph{import} statements, type 
declarations, constant declarations, and template declarations.

\begin{bnflisting}{}
<tslFile>  := <specItem>*
<specItem> := <import>
           |  <typeDecl>
           |  <const>
           |  <template>
\end{bnflisting}

\section{Import statements}

Import statements are used to combine multiple \tsl files into a 
single specification.  They are only allowed in the top-level 
syntactic scope, i.e., they are illegal inside template of type 
declarations.  An import statement consists of the \src{import} 
keyword followed by file path in angle brackets:
\begin{tsllisting}{}
import<ide_dev.tsl>
import<os/ide_tsl2/l4_ide.tsl>
import<../../os/ide_tsl2/ide_class.tsl>
\end{tsllisting}
The \tsl compiler appends this path to each import directory, 
specified via the \src{-I} command line switch, in order, until a  
file with this name is found.

\section{Type declarations}\label{s:r:typedecl}

Type declarations can be placed in the top-level scope or in a
template scope.

\begin{bnflisting}{}
<typeDecl> := "typedef" <typeSpec> <ident>
\end{bnflisting}

\section{Constant declarations}\label{s:r:constant}

Constant declarations can be placed in the top-level scope or in a
template scope.  The value of a constant is an expression that can
be evaluated at compile time.

\begin{bnflisting}{}
<const> := "const" <typeSpec> <ident> "=" <expr>
\end{bnflisting}

Examples:

\begin{tsllisting}{}
typedef struct {bool f1; uint<16> f2;} stype;

const bool     b = true;
const uint<16> u = 16'd5;
const stype    s = stype {.f1 = b, .f2 = u};
\end{tsllisting}


\section{Templates}

Templates must be declared within the top-level scope, i.e., 
nested template declarations are not allowed.  Template 
declaration has the following syntax:
\begin{bnflisting}{}
<template> := "template" <ident> [(<portDeclarations>)]
                  (<templateItem> ";")*
              "endtemplate"
\end{bnflisting}
Here, \src{<portDeclarations>} is a comma-separated list of port 
declarations, described in Section~\ref{s:o:templates}.  A 
template item is one of:
\begin{itemize}
    \item Derive statement
    \item Instance declaration
    \item Type declaration
    \item Constant declaration
    \item Global variable declaration
    \item Init block
    \item Always block
    \item Process
    \item Method
    \item Goal
    \item Wire declaration
\end{itemize}

Derive statements and instance declarations were considered in 
Section~\ref{s:o:templates}.  Type declarations and constant 
declarations were considered in Sections~\ref{s:r:typedecl} and
\ref{s:r:constant}.  We describe the remaining types of template 
items below.

\subsection{Global variable declarations}

Global variables (Section~\ref{s:o:variables}) are declared in the 
template scope.  The declaration syntax is the same as for local 
variables (Section~\ref{s:r:localvar}) with the optional 
\src{export} qualifier that indicates that the variable can be 
accessed from outside the template:

\begin{bnflisting}{}
<gvarDecl> := ["export"] <varDecl>
\end{bnflisting}

Similar to local variable declarations, if the declaration does 
not specify an initial value for the variable, the value for the 
variable is chosen non-deterministically.  The initial value can 
be further constrained by \src{init} blocks, as described below.

\subsection{Init blocks}

An \src{init} block defines a constraint over initial assignment 
of global veriables.
\begin{bnflisting}{}
<initBlock> := "init" <expr>
\end{bnflisting}
The body of an \src{init} block is a boolean expression over 
global variables visible from the current template, including 
variables exported from other templates.  

The following example constrains initial values of 
\src{config\_in\_progress} and \src{sendq\_head} variables:
\begin{tsllisting}{}
template linux_uart_drv(uart_dev dev)
    bool config_in_progress;
    uint<16> sendq_head;
    init (config_in_progress == true) &&
         (sendq_head == 16'd0);
endtemplate
\end{tsllisting}
Note that the same result can be achieved using initial variable 
assignments:
\begin{tsllisting}{}
template linux_uart_drv(uart_dev dev)
    bool config_in_progress = true;
    uint<16> sendq_head = 16'd0;
endtemplate
\end{tsllisting}
In general, however, \src{init} blocks are a more general 
mechanism for constraining initial state, for example the 
following condition cannot be captures using initial variable 
assignments:
\begin{tsllisting}{}
template linux_uart_drv(uart_dev dev)
    init (config_in_progress == true) ||
         (sendq_head == 16'd0);
endtemplate
\end{tsllisting}

A template can contain multiple \src{init} blocks.  In using 
\src{init} blocks, one must make sure that different \src{init} 
blocks do not contradict each other and initial variable 
assignments, leading to an empty initial set, as in the following 
example:
\begin{tsllisting}{}
template linux_uart_drv(uart_dev dev)
    bool config_in_progress = true;
    init config_in_progress == false;
endtemplate
\end{tsllisting}

\subsection{Always blocks}\label{s:r:always}

An always block is an arbitrary statement that is automatically 
prepended to all transitions of the system.  It is intended as a 
low-level mechanism that allows implementing certain behaviours 
that are tricky to achieve without it.  It should be used with 
care and should probably be hidden behind syntactic sugar.

\begin{bnflisting}{}
<always> := "always" <statement>
\end{bnflisting}

\subsection{Processes}

See Section~\ref{s:o:execmodel}.

\begin{bnflisting}{}
<processDecl> := "process" <ident> <statement>
\end{bnflisting}

\subsection{Methods}\label{s:r:method}

A method declaration consists of
\begin{itemize}
    \item optional \src{export} qualifier that indicates whether 
        the method can be invoked from outside its template
    \item method category specifier that labels the method as 
        function, procedure, or task, and in the last case 
        optionally as a controllable or uncontrollable task 
        (Section~\ref{s:o:execmodel})
    \item return type or \src{void} if the method does not return 
        a value
    \item argument list
    \item method body
\end{itemize}

\begin{bnflisting}{}
<methodDecl> := ["export"]               // Exported method?
                <methCateg>              // Method category
                ("void" | <typeSpec>)       // Return type
                <ident> "("                 // Method name
                    [<arg> ("," <arg>)*]    // Arguments
                ")"
                    (                       // Partial body:
                     ["before" <statement>] // preamble
                     ["after" <statement>]  // epilogue
                    ) |
                    <statement>             // complete form

<methCateg> := "function"
             | "procedure"
             | "task" [ "controllable" 
                      | "uncontrollable"]

<arg> := ["out"] <typeSpec> <ident>
\end{bnflisting}

Method arguments are used to pass data both to and from the 
method.  Each individual argument can be declared as an input or 
an output argument, but not both.  Syntactically, input and output 
arguments are distinguished using the \src{out} qualifier.

Both input and output arguments can be read and modified in the 
body of the method; however the initial value of an output 
argument is non-derministic.  The final value of an input argument 
is dropped, while the final value of an output argument is 
propagated to the L-expression (Section~\ref{s:r:lexpr}) passed as 
the actual argument to the method.  Note that for controllable and 
uncontrollable tasks the output value is propagated to the 
argument at the beginning of the next transition performed by the 
process after the implicit pause location following the return 
statement (Section~\ref{s:r:return}).

\subsubsection{Method body}

\comment{The method overriding mechanism described in this section 
is probably too complex and not very useful in practice.  We will 
most likely simplify it by getting rid of \src{before} and \src{after}
constructs.}

Method body declaration can be written in the \emph{complete} or 
\emph{partial} form.  A complete declaration is simply a \tsl 
statement.  A partial declaration consists of an optional preamble 
and an optional epilogue that are intended to execute respectively 
before and after the main body of the method declared in the child 
template.  If neither the preamble not the epilogue are specified, 
then the body of the method remains empty.  

The full method body is constructed by recursively merging the method 
body provided in the method declaration with the overloaded method 
declaration in the parent template (if one exists) using 
Algorithm~\ref{a:method}.  The following examples illustrate the algorithm.


\begin{algorithm}
    \begin{algorithmic}[1]
        \Statex {\bf Input:} Method body $M$, where $M$ is 
        either a complete body $(b)$ or partial body $(p,e)$.
        \Statex {\bf Output:} Method body merged with the parent body (if one exists)
        \Function{fullBody}{$M$}
          \State{\it // Find previous method declaration in one of parent templates}
          \State $M' \gets$ \Call{parentMethod}{$M$}
          \If{$M'=nil$}
            \State {\it // $M$ is not an overloaded method}
            \State {\bf return} $M$
          \EndIf
          \Switch{(M, \Call{fullBody}{$M'$})}
            \State{\it //Parent method has a complete body: override it}
            \Case{$(M, (b'))$}
              \Return{$M$}
            \EndCase
            \State{\it // Both parent and child have partial bodies: merge}
            \State{\it // preambles and epilogues using sequential composition}
            \Case{$((p,e), (p',e'))$} 
              \Return{$((p';p), (e;e'))$}
            \EndCase
            \State{\it // Child has a complete body, parent has partial body: prepend}
            \State{\it // and append parent's preamble and epilogue to the child}
            \Case{$((b), (p',e'))$} 
              \Return{$(p';b;e')$}
            \EndCase
            \Case{\bf otherwise}
              \Return {\bf error}
            \EndCase
          \EndSwitch 
        \EndFunction
    \end{algorithmic}
    \caption{}\label{a:method}
\end{algorithm}

\begin{tsllisting}{}
// global variable
bool x;

// parent declaration
procedure void p(uint<16> arg)
before{ 
    assume(arg != 0);
};
after{
    assert(x);
};

// child declaration
procedure void p(uint<16> arg)
{
    x = (arg > 5);
};

// full child method body generated by the compiler
procedure void p(uint<16> arg)
{
    assume(arg != 0);
    x = (arg > 5);
    assert(x);
};
\end{tsllisting}

\begin{tsllisting}{}
// parent declaration
procedure void p(uint<16> arg)
before{ 
    assume(arg != 0);
};
after{
    assert(x);
};

// child declaration
procedure void p(uint<16> arg)
before{
    x = (arg > 5);
};

// full child method body generated by the compiler
procedure void p(uint<16> arg)
before{
    assume(arg != 0);
    x = (arg > 5);
};
after{
    assert(x);
};
\end{tsllisting}


\subsection{Goals}

\begin{bnflisting}{}
<goalDecl> := "goal" <ident> "=" <expr>
\end{bnflisting}

See Section~\ref{s:o:correctness}.

\subsection{Wire declarations}

Wire declaration (Section~\ref{s:o:variables}) consists of wire 
type, wire name, and optional wire expression.  

\begin{bnflisting}{}
<wire> := ["export"] "wire" <typeSpec> <ident> ["=" <expr>]
\end{bnflisting}

The wire expression is defined over global variables and wires of 
the current template and other templates accessible via 
hierarchical identifiers.  However, circular dependencies among 
wire expressions are forbidden.  If the wire expression is 
omitted, the wire is a \emph{pure wire} that must be re-defined in 
child templates.  In merging template with its parents, the \tsl 
compiler picks the last wire declaration in the inheritance 
hierarchy.

\end{document}
